### 2014-11-10 heat-transfer.tex
Add locking section.

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. XXX rewrite The program stops when the new heat values of every node are
$\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize between iterations.

@@

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.
\end{Verbatim}
  \scap{code:ht}{Coordination code for the Heat Transfer program. We updated the first rule
     to increase the priority of neighbor nodes.  The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
\end{topfig}
\normalsize

@@

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old)
fabs(New - Old) <= epsilon
\underline[cpu-id(A, B, C)],
\underline[cpu-id(A, A, C)]
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.

>>>>

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The program stops when the new heat values of every node are
$\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize between iterations.

@@

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon
   -o {B | !edge(A, B) |
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
   -o {B | !edge(A, B) |
         new-neighbor-heat(B, A, New)}.
\end{Verbatim}
  \scap{code:ht}{Coordination code for the Heat Transfer program. We updated the first rule
     to increase the priority of neighbor nodes.  The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
\end{topfig}
\normalsize

@@

\begin{topfig}
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old)
fabs(New - Old) <= epsilon
\underline[cpu-id(A, B, C)],
\underline[cpu-id(A, A, C)]
   -o {B, W | !edge(A, B) |
         new-neighbor-heat(B, A, New)}.

### 2014-11-10 implementation.tex
fix ie
Add locking section.

\subsubsection{Coordination directives}

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediatelly used by the machine, there is no need to store them in the
database, therefore avoiding unnecessary allocations and deallocations.

@@

\subsubsection{Nodes}

A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available facts and potential candidate rules. We need
a separate indexing structure per node since rules run local to each node.

@@

\subsubsection{Threads}

The main loop for each thread is to get an active node from one of its
local queues and process all applicable rules for that node generating
new facts which can result in new facts for the current node or other
nodes.  The thread will continue processing facts for the current node
until there are no new facts generated for it.  It will then pick
another active node.  If its queues are all empty it will try to steal
work (\ie nodes) from another thread.

@@

\subsubsection{Communication}

@@

\subsubsection{Garbage collection}

@@

\subsubsection{Coordination overhead}

>>>>

\paragraph{Coordination directives}
Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediatelly used by the machine, there is no need to store them in the
database, therefore avoiding unnecessary allocations and deallocations.

@@

\paragraph{Nodes} 
A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available facts and potential candidate rules. We need
a separate indexing structure per node since rules run local to each node.

@@

\paragraph{Threads}

The main loop for each thread is to get an active node from one of its
local queues and process all applicable rules for that node generating
new facts which can result in new facts for the current node or other
nodes.  The thread will continue processing facts for the current node
until there are no new facts generated for it.  It will then pick
another active node.  If its queues are all empty it will try to steal
work (i.e., nodes) from another thread.

@@

\paragraph{Communication}

@@

\paragraph{Garbage collection}

@@

\paragraph{Coordination overhead}

### 2014-11-10 locks.tex
Merge HT locks plots into one.
Update figures with SSSP.
Add locking section.

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks-local-only.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/minmax-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/sssp-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/bp.png}}
   \end{center}
\end{topfig}

>>>>

\paragraph{Locking} The CLM virtual machine uses spinlocks to protect shared data structures between
threads. As described in Section~\ref{sec:implementation}, coordination
operations require synchronization since they need to change node information
and sometimes move nodes between the standard queue and the regular queue.
In order to measure the locking costs of coordination, we measured the following
statistics: \emph{basic locks}, the number of locks acquired that were already
present in the original LM virtual machine and \emph{coordination locks}, the
number of locks acquired for performing coordination operations. We also
measured how often a \texttt{lock()} operation fails on the first attempt, but
such statistics are not represented in the figures below since the values are
too small.

The first program we measured was SSSP, represented in
Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
Coordinated versions, we have the Buffered version where we
optimized the virtual machine to buffer coordination actions before applying
them. The idea is to avoid applying multiple \texttt{set-priority} operations to
the same node during a single node execution. We thus gather the higher priority
values and then perform coordination after the node has executed.
Although the number of locks is significantly reduced, we did not see a
similar performance improvement, since the Buffered version is only
5\% faster than the Coordinated version.

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/sssp-locks.png}}
   \end{center}
  \scap{locks:SSSP}{Locking statistics for SSSP.
     \texttt{Coordinated} refers to the program in
     Fig.~\ref{results:sssp_uspowergrid}, while \texttt{Buffered} is a more
     optimized version of the same program where coordination operations are
     buffered before application.}
\end{topfig}

In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
regular version, the coordinated version and the local-only version as in
Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
coordination and then a further reduction in the local-only version because less
facts are exchanged between threads.

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
   \end{center}
  \scap{locks:HT}{Locking statistics for HT. Coordination reduces
     the overall number of locks used, while further optimization for locality
     reduces the locks even further.}
\end{topfig}

Finally, in Fig.~\ref{locks:LBP} we present lock statistics for LBP and SBP. LBP
is represented as the Regular bar, while SBP is represented as the Coordinated
bar. We notice that LBP has quite a lot of locking for 1 thread since LBP
is an asynchronous algorithm and tends to be a slow sequential algorithm,
requiring more derivations when using 1 thread. The numbers go down as more
threads are used. For SBP, there is a consistent amount of locking without
little variations between the number of threads used. This is because threads
interactly very little with each other and the amount of work to do is the same
for all threads.

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/bp.png}}
   \end{center}
  \scap{locks:LBP}{Locking statistics for LBP and SBP. Locking is constant for
     SBP because threads do not interact with each other.}
\end{topfig}

These experiments indicate that handling coordination has a small cost that is
greatly offset by improved performance. In general, coordination reduces the
amount of locking required in the overall program because the number of facts
needed to be derived is reduced. Even in cases where the amount of locking
increases, it is still better to use coordination due to less synchronization
between threads.
